---
title: "Quantifying Clonal Bias"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Quantifying Clonal Bias}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.9em;
}
</style>
 
```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)

suppressMessages(library(scRepertoire))
suppressMessages(library(Seurat))
data("contig_list") 
combined.TCR <- combineTCR(contig_list, 
                           samples = c("P17B", "P17L", "P18B", "P18L", 
                                            "P19B","P19L", "P20B", "P20L"))

scRep_example <- readRDS("scRep_example_full.rds")

scRep_example <- combineExpression(combined.TCR, 
                                   scRep_example, 
                                   cloneCall="gene", 
                                   group.by = "sample")

#Adding patient information
scRep_example$Patient <- substr(scRep_example$orig.ident, 1,3)

#Adding type information
scRep_example$Type <- substr(scRep_example$orig.ident, 4,4)

#Defining colors
colorblind_vector <- hcl.colors(n=7, palette = "inferno", fixup = TRUE)
```

## StartracDiversity

From the excellent work by [Zhang et al. (2018, Nature)](https://www.nature.com/articles/s41586-018-0694-x), the authors introduced new methods for looking at clones by cellular origins and cluster identification. We strongly recommend you read and cite their publication when using this function.

To use the ```StartracDiversity()``` function, you need the output of the ```combineExpression()``` function and a column in your metadata that specifies the tissue of origin. 

### Indices Output from `StartracDiversity()`

*  `expa` - Clonal Expansion. Measures the degree of clonal proliferation within a given cell cluster. It is calculated as `1 - normalized Shannon entropy`, where a higher value indicates that a few clones dominate the cluster.  
*  `migr` - Cross-tissue Migration. Quantifies the movement of clonal T cells between different tissues, as defined by the `type` parameter. This index is based on the entropy of a single clonotype's distribution across the specified tissues.  
*  `tran` - State Transition. Measures the developmental transition of T cell clones between different functional clusters. This index is calculated from the entropy of a single clonotype's distribution across the cell clusters identified in the data.

Key Parameters for `StartracDiversity()`

* `type`: The variable in the metadata that provides tissue type.
* `group.by`: A column header in the metadata to group the analysis by (e.g., "sample", "treatment"). 

By default, `StartracDiversity()` will calculate all three indices (`expa`, `migr`, and `tran`) for each cluster and group you define. This provides a comprehensive overview of the clonal dynamics in your dataset. The output is a three-paneled plot, with each panel representing one index.

In the example data, `type` corresponds to the "Type" column, which includes "P" (peripheral blood) and "L" (lung) classifiers. The analysis is grouped by the "Patient" column.

```{r}
# Calculate and plot all three STARTRAC indices
StartracDiversity(scRep_example, 
                  type = "Type", 
                  group.by = "Patient")
```

### Calculating a Single Index

If you're only interested in one aspect of clonal dynamics, you can specify it using the `index` parameter. For example, to only calculate and plot clonal expansion:

```{r}
# Calculate and plot only the clonal expansion index
StartracDiversity(scRep_example, 
                  type = "Type", 
                  group.by = "Patient",
                  index = "expa")
```

### Pairwise Migration Analysis

Another feature of `StartracDiversity()` is the ability to perform pairwise comparisons. To specifically quantify the migration between two tissues (e.g., Lung vs. Periphery), set `index = "migr"` and tell the function which column to use for the comparison with `pairwise = "Type"`.

```{r}
# # Calculate pairwise migration between tissues
StartracDiversity(scRep_example,
                  type = "Type",
                  group.by = "Patient",
                  index = "migr",
                  pairwise = "Type")
```

### Pairwise Transition Analysis

Setting `index = "tran"` and `pairwise = "cluster"` does the same thing across clusters. The function walks every unordered pair of clusters, subsets the cells to that pair, and recomputes the transition index inside that two-cluster subspace.

```{r}
StartracDiversity(scRep_example,
                  type = "Type",
                  group.by = "Patient",
                  index = "tran",
                  pairwise = "cluster",
                  export.table = TRUE)[1:6, ]
```

Reading the table:

* `group` - the `group.by` level the row was calculated in.
* `cluster` - the cluster the score is anchored to, meaning the one supplying the cell weights.
* `partner` - the other cluster in the pair.
* `comparison` - the pair itself. Labels are always sorted, so the same pair reads the same way in every group.
* `value` - the cell-weighted mean clonal entropy over the pair. With only two categories it runs from 0 to 1. A `0` means the clones making up the anchor cluster are exclusive to it. A `1` means every clone in the anchor is split evenly with the partner.

Two rows exist for each pair because each cluster takes a turn as the anchor. They usually differ, and that asymmetry comes from the weighting rather than from any direction of travel. A pair labeled `1 vs 3` is not a claim that cells moved from cluster 1 to cluster 3.

This matters if you are grouping by time point. STARTRAC describes clonal sharing as it stands, so it cannot order that sharing in time on its own. To ask whether a clone moved from a cluster at t1 into a different cluster at t2, track clone identity across the time points directly, for example with `alluvialClones()` or by tabulating each clone's cluster membership per time point.

## clonalBias

A new metric proposed by [Massimo et al](https://pubmed.ncbi.nlm.nih.gov/35829695/), ```clonalBias()```, like STARTRAC, is a clonal metric that seeks to quantify how individual clones are skewed towards a specific cellular compartment or cluster. A clone bias of `1` indicates that a clone is composed of cells from a single compartment or cluster, while a clone bias of `0` matches the background subtype distribution. Please read and cite the linked manuscript if using `clonalBias()`

Key Parameter(s) for `clonalBias()`

* `group.by`: A column header in the metadata that bias will be based on.
* `split.by`: The variable to use for calculating the baseline frequencies (e.g., "Type" for lung vs peripheral blood comparison)
* `n.boots`: Number of bootstraps to downsample.
* `min.expand`: Clone frequency cut-off for the purpose of comparison (default = 10).

Here we calculate and plot clonal bias using `aa` clone calls, splitting by "Patient" and grouping by "seurat_clusters", with a minimum expansion of 5 and 10 bootstraps:

```{r, message = FALSE,}
clonalBias(scRep_example, 
           cloneCall = "aa", 
           split.by = "Patient", 
           group.by = "seurat_clusters",
           n.boots = 10,
           min.expand =5)
```

## Related Articles

- [Visualizations for Single-Cell Objects](SC_Visualizations.html) - Alluvial plots, chord diagrams, and clonal network overlays.
- [Combining Clones and Single-Cell Objects](Attaching_SC.html) - Attach clonal data with `combineExpression()`.
- [Comparing Clonal Diversity and Overlap](Clonal_Diversity.html) - Diversity metrics and repertoire overlap.
