---
title: "Bridging Python TCR Tools with immLynx"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
theme: united
df_print: kable
vignette: >
  %\VignetteIndexEntry{Bridging Python TCR Tools with immLynx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
p.caption {
  font-size: 0.9em;
}
</style>

```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  eval = FALSE
)
```

# Getting Started

**immLynx** is a comprehensive R package that bridges popular Python-based immune repertoire analysis tools and Hugging Face protein language models into the R environment. It provides unified interfaces for:

- TCR distance calculations ([tcrdist3](https://github.com/kmayerb/tcrdist3))
- Sequence generation probability ([OLGA](https://github.com/statbiophys/OLGA))
- Selection inference ([soNNia](https://github.com/statbiophys/soNNia))
- TCR clustering ([clusTCR](https://github.com/svalkiers/clusTCR))
- Protein language model embeddings ([ESM-2](https://github.com/facebookresearch/esm))
- Metaclone discovery ([metaclonotypist](https://github.com/emerson-hammer/metaclonotypist))
- Near-neighbor search by symmetric deletion lookup ([pyrepseq](https://github.com/andim/pyrepseq))
- Deep generative featurization ([DeepTCR](https://github.com/sidhomj/DeepTCR))
- Clonal expansion prediction from gene expression ([scXpand](https://github.com/scXpand/scXpand))
- Export to the Python single-cell ecosystem ([scanpy](https://github.com/scverse/scanpy) and [scirpy](https://github.com/scverse/scirpy))

immLynx is fully compatible with the [scRepertoire](https://github.com/BorchLab/scRepertoire) and [immApex](https://github.com/BorchLab/immApex) ecosystem for single-cell immune repertoire analysis.

More information is available at the [immLynx GitHub Repo](https://github.com/BorchLab/immLynx).

## Installation

```{r, eval = FALSE}
devtools::install_github("BorchLab/immLynx")
```

### Python Dependencies

immLynx uses [basilisk](https://bioconductor.org/packages/basilisk/) to manage Python dependencies automatically. The first time you use functions that call Python, basilisk will create an isolated conda environment with all required packages. This may take a few minutes on first run.

A few tools pin dependency versions that cannot coexist with that main environment, so they get their own. `exportToScanpy()` uses a scanpy/scirpy environment, `runDeepTCR()` uses a TensorFlow environment, and `runScXpand()` uses a PyTorch environment. Each is built on first use of the function that needs it, so you only pay for the ones you actually call.

## The Data Set

To demonstrate the analysis functions in immLynx, we will use the scRepertoire example data derived from T cells of 4 patients with acute respiratory distress. The data contains paired peripheral-blood (B) and bronchoalveolar lavage (L) samples across 8 distinct runs from the 10x Cell Ranger pipeline.

## Loading Libraries and Data

```{r}
suppressMessages(library(scRepertoire))
suppressMessages(library(immLynx))
suppressMessages(library(Seurat))
suppressMessages(library(ggplot2))
suppressMessages(library(viridis))
suppressMessages(library(dplyr))

data("contig_list")
combined.TCR <- combineTCR(contig_list,
                           samples = c("P17B", "P17L", "P18B", "P18L",
                                            "P19B","P19L", "P20B", "P20L"))

scRep_example <- readRDS("scRep_example_full.rds")

scRep_example <- combineExpression(combined.TCR,
                                   scRep_example,
                                   cloneCall="aa",
                                   group.by = "sample",
                                   proportion = TRUE)

#Filtering for single-cells with TCRs
scRep_example <- subset(scRep_example,
                        cells = colnames(scRep_example)[!is.na(scRep_example$CTaa)])
```

# Extracting and Validating TCR Data

Before running analyses, you can extract and validate TCR data from the single-cell object:

```{r}
# Extract beta chain data
tcr_data <- extractTCRdata(scRep_example, chains = "TRB")
head(tcr_data)

# Extract both chains in wide format
tcr_wide <- extractTCRdata(scRep_example, chains = "both", format = "wide")

# Validate the data
validation <- validateTCRdata(tcr_data)
print(validation)
```

# Summarizing the TCR Repertoire

Get a comprehensive overview of the TCR repertoire:

```{r}
summary <- summarizeTCRrepertoire(scRep_example, chains = "TRB")
print(summary)
```

# Analysis Functions

## TCR Clustering with clusTCR

Cluster TCRs based on sequence similarity using the clusTCR algorithm. The MCL (Markov Clustering) algorithm identifies groups of TCRs with similar CDR3 sequences:

```{r}
scRep_example <- runClustTCR(
  scRep_example,
  chains = "TRB",
  method = "mcl",
  inflation = 2.0
)

table(scRep_example$clustcr_TRB)
```

### Comparing Clustering Parameters

Different inflation parameters control the granularity of clustering:

```{r}
# MCL clustering with different inflation parameters
scRep_example <- runClustTCR(scRep_example, method = "mcl", inflation = 2.0,
                            column_prefix = "mcl_2")
scRep_example <- runClustTCR(scRep_example, method = "mcl", inflation = 3.0,
                            column_prefix = "mcl_3")

# DBSCAN clustering
scRep_example <- runClustTCR(scRep_example, method = "dbscan", eps = 0.5,
                             column_prefix = "dbscan")

# Compare number of clusters
cat("MCL (inflation=2):", length(unique(na.omit(scRep_example$mcl_2_TRB))), "clusters\n")
cat("MCL (inflation=3):", length(unique(na.omit(scRep_example$mcl_3_TRB))), "clusters\n")
cat("DBSCAN:", length(unique(na.omit(scRep_example$dbscan_TRB))), "clusters\n")
```

## Near-Neighbor Search with symdel

Symmetric deletion lookup finds every pair of CDR3 sequences within a given edit distance without building a full pairwise distance matrix. It indexes each sequence by its set of single-character deletions, so two sequences within one edit share a deletion variant and collide in the index. Cost scales with the number of sequences rather than with their square.

By default `runSymdelNeighbors()` adds a `symdel_degree` column giving the number of neighbors each cell's CDR3 has. High degree marks sequences sitting in dense similarity neighborhoods, the signature of convergent recombination and antigen-driven expansion:

```{r}
scRep_example <- runSymdelNeighbors(
  scRep_example,
  chains = "TRB",
  max_edits = 1
)

table(scRep_example$symdel_degree, useNA = "ifany")
```

Setting `return_object = FALSE` returns the neighbor pairs themselves, which is the more useful form when building a similarity graph:

```{r}
edges <- runSymdelNeighbors(
  scRep_example,
  chains = "TRB",
  max_edits = 2,
  return_object = FALSE
)

head(edges)
```

The edge list feeds directly into a graph representation of the repertoire:

```{r}
library(igraph)

g <- graph_from_data_frame(edges[, c("from_seq", "to_seq")], directed = FALSE)
components <- components(g)

cat("Connected components:", components$no, "\n")
cat("Largest component:", max(components$csize), "sequences\n")
```

This is the same algorithm benchmarked in the [XT-neighbor](https://github.com/heartnetkung/XT-neighbor) work. immLynx uses the CPU implementation from pyrepseq, so no GPU is required.

## TCR Distance Calculations with tcrdist3

Calculate pairwise TCR distances using the tcrdist3 framework:

```{r}
dist_results <- runTCRdist(
  scRep_example,
  chains = "beta",
  organism = "human"
)

dim(dist_results$distances$pw_beta)
```

The distance matrix can also be used for hierarchical clustering or other downstream analyses:

```{r}
d <- as.dist(dist_results$distances$pw_beta)
hc <- hclust(d, method = "complete")
clusters <- cutree(hc, k = 50)
```

## Generation Probability with OLGA

Calculate the generation probability (Pgen) for TCR sequences. Pgen reflects the probability of a given TCR sequence being generated by V(D)J recombination:

```{r}
scRep_example <- runOLGA(
  scRep_example,
  chains = "TRB",
  model = "humanTRB"
)

hist(log10(scRep_example$olga_pgen_TRB),
     breaks = 50,
     main = "Distribution of TCR Generation Probabilities",
     xlab = "log10(Pgen)")
```

### Comparing Pgen Across Sample Types

Cells with low Pgen may be under stronger selection because they are rare in the naive repertoire:

```{r}
ggplot(scRep_example@meta.data, aes(x = Type, y = olga_pgen_log10_TRB)) +
  geom_boxplot() +
  labs(title = "TCR Generation Probability by Sample Type",
       x = "Sample Type",
       y = "log10(Pgen)") +
  theme_classic()
```

### Generating Random TCR Sequences

OLGA can also generate random TCR sequences from the recombination model:

```{r}
random_seqs <- generateOLGA(n = 100, model = "humanTRB")
head(random_seqs)
```

## Protein Language Model Embeddings

Generate TCR embeddings using ESM-2 protein language models. These embeddings capture biochemical and structural properties of the CDR3 sequences:

```{r}
scRep_example <- runEmbeddings(
  scRep_example,
  chains = "TRB",
  model_name = "facebook/esm2_t12_35M_UR50D",
  pool = "mean"
)

scRep_example <- RunUMAP(scRep_example,
                          reduction = "tcr_esm",
                          dims = 1:30,
                          reduction.name = "tcr_umap",
                          reduction.key = "tcrUMAP_")

DimPlot(scRep_example, reduction = "tcr_umap")
```

### Comparing ESM-2 Model Sizes

ESM-2 comes in different sizes. Larger models may capture more nuanced biochemical properties but require more computational resources:

```{r}
# Small model (35M parameters) - fast, lower quality
scRep_example <- runEmbeddings(
  scRep_example,
  model_name = "facebook/esm2_t12_35M_UR50D",
  reduction_name = "esm_small"
)

# Medium model (650M parameters) - balanced
scRep_example <- runEmbeddings(
  scRep_example,
  model_name = "facebook/esm2_t33_650M_UR50D",
  reduction_name = "esm_medium"
)

# Compare embeddings via UMAP
scRep_example <- RunUMAP(scRep_example, reduction = "esm_small", dims = 1:30,
                          reduction.name = "umap_small", reduction.key = "umapSmall_")
scRep_example <- RunUMAP(scRep_example, reduction = "esm_medium", dims = 1:30,
                          reduction.name = "umap_medium", reduction.key = "umapMed_")

p1 <- DimPlot(scRep_example, reduction = "umap_small") + ggtitle("ESM-2 35M")
p2 <- DimPlot(scRep_example, reduction = "umap_medium") + ggtitle("ESM-2 650M")
p1 + p2
```

### Embedding Both Chains

For paired alpha-beta TCR analysis:

```{r}
scRep_example <- runEmbeddings(
  scRep_example,
  chains = "both",
  reduction_name = "tcr_paired"
)

scRep_example <- RunUMAP(scRep_example,
                          reduction = "tcr_paired",
                          dims = 1:30,
                          reduction.name = "paired_umap",
                          reduction.key = "pairedUMAP_")

DimPlot(scRep_example, reduction = "paired_umap")
```

### Custom Embedding Workflow

For advanced users who need fine-grained control over the embedding process, immLynx exposes the lower-level functions used internally by `runEmbeddings()`:

```{r}
# Step 1: Load model and tokenizer
model_info <- huggingModel(model_name = "facebook/esm2_t12_35M_UR50D")

# Step 2: Extract CDR3 sequences
sequences <- scRep_example$CTaa
sequences <- gsub("_.*", "", sequences)  # Get first chain
sequences <- unique(na.omit(sequences))

# Step 3: Tokenize sequences
tokenized <- tokenizeSequences(
  tokenizer = model_info$tokenizer,
  aa_sequences = sequences,
  padding = TRUE,
  truncation = TRUE
)

# Step 4: Generate embeddings with custom settings
embeddings <- proteinEmbeddings(
  model = model_info$model,
  tokenized.batch = tokenized,
  pool = "mean",
  prefer_device = "auto"
)

dim(embeddings)
```

## Deep Generative Featurization with DeepTCR

DeepTCR learns a latent representation of CDR3 sequences with a variational autoencoder. Where ESM-2 brings general protein knowledge learned from UniProt, the DeepTCR VAE trains on the repertoire in front of it, so its features describe the structure of that specific dataset. The two are complementary rather than interchangeable.

DeepTCR prunes uninformative latent dimensions during training, so the result may be narrower than `latent_dim` requests:

```{r}
scRep_example <- runDeepTCR(
  scRep_example,
  chains = "TRB",
  latent_dim = 64,
  verbose = FALSE
)

dim(Embeddings(scRep_example, "tcr_deeptcr"))
```

Passing `use_genes = TRUE` includes V and J gene usage alongside the CDR3 sequence, which helps when gene usage carries specificity signal the CDR3 alone does not:

```{r}
scRep_example <- runDeepTCR(
  scRep_example,
  chains = "TRB",
  use_genes = TRUE,
  reduction_name = "tcr_deeptcr_genes"
)
```

### Comparing DeepTCR and ESM-2 Representations

Because both write to a dimensional reduction, they live in the same object and can be projected side by side:

```{r}
scRep_example <- runEmbeddings(scRep_example, chains = "TRB",
                               reduction_name = "tcr_esm")
scRep_example <- runDeepTCR(scRep_example, chains = "TRB",
                            latent_dim = 64, verbose = FALSE)

scRep_example <- RunUMAP(scRep_example, reduction = "tcr_esm",
                         dims = 1:64, reduction.name = "umap_esm")
scRep_example <- RunUMAP(scRep_example, reduction = "tcr_deeptcr",
                         dims = 1:64, reduction.name = "umap_deeptcr")

DimPlot(scRep_example, reduction = "umap_esm", group.by = "clustcr_TRB") +
  ggtitle("ESM-2") + NoLegend()
DimPlot(scRep_example, reduction = "umap_deeptcr", group.by = "clustcr_TRB") +
  ggtitle("DeepTCR VAE") + NoLegend()
```

The VAE needs enough unique sequences to learn anything meaningful. On a few hundred sequences the latent space is dominated by noise, and the reconstruction accuracy printed during training is the quickest way to see it.

## Clonal Expansion Prediction with scXpand

scXpand predicts T-cell clonal expansion from gene expression alone using pretrained pan-cancer models. Unlike the other wrapped tools it needs no receptor sequences, so it works on datasets with no paired TCR sequencing.

List the available models without building the Python environment:

```{r}
listScXpandModels()
```

Run inference with the default pan-cancer autoencoder. This writes a predicted probability and a thresholded call back to cell metadata:

```{r}
scRep_example <- runScXpand(
  scRep_example,
  model_name = "pan_cancer_autoencoder"
)

hist(scRep_example$scXpand_expansion_prob,
     breaks = 50,
     main = "Predicted Expansion Probability",
     xlab = "scXpand_expansion_prob")

table(scRep_example$scXpand_expansion_pred, useNA = "ifany")
```

When scRepertoire clone calls are present, `runScXpand()` also derives observed clone sizes and an expansion label using scXpand's 1.5x-median rule. That lets a gene-expression-only prediction be scored against the observed repertoire:

```{r}
df <- data.frame(
  prob  = scRep_example$scXpand_expansion_prob,
  truth = scRep_example$scXpand_expansion_truth
)
df <- df[!is.na(df$truth), ]

boxplot(prob ~ truth, data = df,
        main = "Predicted probability vs observed expansion",
        xlab = "Observed expansion", ylab = "scXpand_expansion_prob")
```

Note the model is indexed by Ensembl IDs. If your object uses gene symbols, `runScXpand()` maps them through `org.Hs.eg.db` and reports ambiguous symbols and collapsed duplicates rather than resolving them silently, because scXpand zero-fills genes it cannot find.

## Metaclone Discovery with Metaclonotypist

Identify TCR metaclones - groups of related T cell receptors that may recognize similar antigens:

```{r}
scRep_example <- runMetaclonotypist(
  scRep_example,
  chains = "beta",
  method = "tcrdist",
  max_edits = 2,
  max_dist = 20
)

table(scRep_example$metaclone)
```

## HLA Association Analysis

If HLA typing data is available, you can test associations between metaclone membership and HLA alleles using Fisher's exact test with FDR correction:

```{r}
# First run metaclonotypist
metaclones <- runMetaclonotypist(scRep_example, return_seurat = FALSE)

# Create mock HLA data (replace with real HLA typing data)
hla_data <- data.frame(
  barcode = metaclones$barcode,
  HLA_A_01_01 = sample(c(TRUE, FALSE), nrow(metaclones), replace = TRUE),
  HLA_A_02_01 = sample(c(TRUE, FALSE), nrow(metaclones), replace = TRUE),
  HLA_B_07_02 = sample(c(TRUE, FALSE), nrow(metaclones), replace = TRUE)
)

# Test associations
hla_results <- runHLAassociation(metaclones, hla_data)
head(hla_results)
```

## Selection Inference with soNNia

Infer selection pressures on TCRs using the soNNia framework. This requires a background set of sequences for comparison:

```{r}
# Generate background sequences
background <- generateOLGA(n = 10000, model = "humanTRB")
write.csv(background, "background.csv", row.names = FALSE)

# Run soNNia
scRep_example <- runSoNNia(
  scRep_example,
  chains = "TRB",
  background_file = "background.csv"
)
```

# Combined Analysis Workflow

Here is a complete workflow combining multiple immLynx analyses with the scRepertoire example data:

```{r}
library(immLynx)
library(Seurat)
library(ggplot2)

# Step 1: Cluster TCRs
scRep_example <- runClustTCR(
  scRep_example,
  chains = "TRB",
  method = "mcl"
)

# Step 2: Calculate generation probability
scRep_example <- runOLGA(
  scRep_example,
  chains = "TRB"
)

# Step 3: Generate embeddings
scRep_example <- runEmbeddings(
  scRep_example,
  chains = "TRB"
)

# Step 4: Visualize embeddings
scRep_example <- RunUMAP(
  scRep_example,
  reduction = "tcr_esm",
  dims = 1:30,
  reduction.name = "tcr_umap",
  reduction.key = "tcrUMAP_"
)

# Plot by cluster
plot1 <- DimPlot(scRep_example,
                 reduction = "tcr_umap",
                 group.by = "clustcr_TRB") + NoLegend()

# Plot by Pgen
plot2 <- FeaturePlot(scRep_example,
                     reduction = "tcr_umap",
                     features = "olga_pgen_log10_TRB") +
         scale_color_viridis()

plot1 + plot2
```

# Integration with scRepertoire Clonotypes

immLynx clustering results can be compared directly with scRepertoire clonotype assignments:

```{r}
scRep_example <- runClustTCR(scRep_example, chains = "TRB")

comparison <- table(
  clustcr = scRep_example$clustcr_TRB,
  clonotype = scRep_example$CTstrict
)

cat("Number of clustcr clusters:", nrow(comparison), "\n")
cat("Number of unique clonotypes:", ncol(comparison), "\n")
```

# Handing Off to Python: scanpy and scirpy

`exportToScanpy()` writes the object out for the Python single-cell ecosystem. It accepts a `SingleCellExperiment` or a `Seurat` object and handles the conversion internally.

Writing gene expression alone produces an H5AD file plus an AIRR rearrangement TSV sidecar:

```{r}
res <- exportToScanpy(
  scRep_example,
  output_dir = "scanpy_export",
  format     = "h5ad",
  write_airr = TRUE
)

res$h5ad
res$airr
```

Writing `h5mu` instead assembles gene expression and receptor data into a single MuData file that scirpy can read directly, with chains already indexed:

```{r}
res <- exportToScanpy(
  scRep_example,
  output_dir  = "scanpy_export",
  format      = "h5mu",
  chains      = "both",
  reductions  = c("tcr_esm", "tcr_deeptcr"),
  obs_columns = c("clustcr_TRB", "symdel_degree")
)
```

On the Python side that file loads with no further preparation:

```{python, eval = FALSE}
import muon as mu
import scirpy as ir

mdata = mu.read_h5mu("scanpy_export/scRep_example.h5mu")
ir.tl.chain_qc(mdata)
```

BCR loci (`IGH`, `IGK`, `IGL`) are mapped to the heavy and light chains that `immApex::getIR()` expects and disambiguated by V-gene prefix, so B-cell data exports the same way.

# Exporting Results

Export TCR data for use with external tools:

```{r}
# Export TCR data in tcrdist3 format
tcr_data <- extractTCRdata(scRep_example, chains = "both", format = "wide")
tcrdist_format <- convertToTcrdist(tcr_data, chains = "both")
write.csv(tcrdist_format, "tcr_for_tcrdist.csv", row.names = FALSE)

# Export cluster assignments
clusters <- data.frame(
  barcode = colnames(scRep_example),
  clustcr = scRep_example$clustcr_TRB
)
write.csv(clusters, "cluster_assignments.csv", row.names = FALSE)

# Export embeddings
embeddings <- Embeddings(scRep_example, "tcr_esm")
write.csv(embeddings, "tcr_embeddings.csv")

# Export the neighbor edge list
edges <- runSymdelNeighbors(scRep_example, chains = "TRB",
                            return_object = FALSE)
write.csv(edges, "symdel_edges.csv", row.names = FALSE)
```

# Working with Large Datasets

For large datasets, consider these strategies to manage memory and computation:

```{r}
# 1. Use smaller chunk sizes for embeddings
scRep_example <- runEmbeddings(
  scRep_example,
  chunk_size = 16,  # Smaller chunks for memory
  pool = "mean"
)

# 2. Sample for distance calculations
# For very large datasets, calculate distances on a subset
sample_cells <- sample(colnames(scRep_example), 5000)
subset_obj <- subset(scRep_example, cells = sample_cells)
dist_results <- runTCRdist(subset_obj)

# 3. Use near-neighbor search instead of a distance matrix
# runSymdelNeighbors() avoids the quadratic step entirely and returns a
# sparse edge list, so no subsampling is needed
edges <- runSymdelNeighbors(
  scRep_example,
  chains = "TRB",
  max_edits = 1,
  n_cpu = 4,
  return_object = FALSE
)
```

When the question is "which sequences are similar" rather than "how far apart is every pair", the near-neighbor search is the practical option past a few tens of thousands of unique sequences. For reference, 200,000 sequences at one edit takes roughly 3 seconds on a single core.
